Perturbing SU($n$)-symmetric interactions

### 2020-01-26 collective_sun/sun_perturbation.tex
some improvements to notation in the two-body eigenvalue problem in the appendices

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \p{2h_{pq}-h_p-h_q} w_{pq}
  + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
where $h_p\equiv\sum_q h_{pq}$.  If the interaction Hamiltonian $H_0$
is translationally invariant, then the couplings $h_{pq}$ depend on
$p$ and $q$ only through the separation $\pm\p{p-q}$.  In this case,
all $h_p=h_0$ are equal, and \eqref{eq:cond_2_ref_start} can be
written in the form
\begin{align}
  2 \m h*\m w + \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h}
  = \p{\Delta + 2h_0} \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$ and $\m w$ are matrices of all $h_{pq}$ and $w_{pq}$;
$\m h*\m w$ denotes an element-wise (Kronecker) product of $\m h$ and
$\m w$, i.e.~with matrix elements $h_{pq}w_{pq}$; and
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices.  In order to find solutions to \eqref{eq:cond_2_ref}, we we
expand the matrices $\m h,\m w$ in a basis that the respects
translational symmetry of $\m h$.  To this end, for all lattice
displacements $d\in\ZZ_L^D$ and wavenumbers $k\in\ZZ_L^D\times2\pi/L$
we define the {\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms of the cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
expand
\begin{align}
  \m h = \sum_{p,q} h_{pq} \op{p}{q}
  = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
\end{align}
and write the condition in \eqref{eq:cond_2_ref} as a new set of
conditions
\begin{align}
  \sum_{d\ne0} 2 h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  = \p{\Delta + 2 h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{d\ne0} 2 h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  = \p{\Delta + 2 h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{d\ne0} 2 h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k
  = \p{\Delta + 2 h_0} \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a fully symmetric
state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a fixed-energy
perturbation $\V_2^\Delta$ is then performed by projecting each
$k$-component of $\v{\m w}$ onto the corresponding $\Delta$-eigenspace
of $\ut{\check{\m h}}_k$.  If the perturbation $\V_2$ is
translationally invariant, then the $k$-components of $\v{\m w}$ for
all $k\ne 0$ all vanish, so we only need to consider the case of
$k=0$.

@@

Here we solve the single- and two-body eigenvalue problems in the case
of a fully collective spin model with uniform couplings:
$h_{pq}=-1+\delta_{pq}$.  In this case, the single-body excitation
energies $\Delta_k$ are
\begin{align}
  \Delta_k
  = - \sum_{d\ne 0} \sp{\cos\p{d\c k}-1}
  = N \p{1-\delta_{k,0}}.
\end{align}
In order to solve the two-body eigenvalue problem, we expand
\begin{align}
  \ut{\check{\m h}}_k
  = - \sum_{d\ne0} 2 \op{d}
  - \sum_{\substack{c,d\ne0\\c\ne d}} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c}
  = - \sum_{c,d\ne0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
\end{align}
and define the vectors
\begin{align}
  \ket{\ut{p}}
  \equiv \f1{\sqrt{N-1}} \sum_{d\ne 0} e^{i p\c d}\ket{d},
\end{align}
in terms of which
\begin{align}
  \ut{\check{\m h}}_k
  = -\p{N-1} \p{\op{\ut{0}} + \op{\ut{k}}}.
\end{align}
The vectors $\ket{\ut{p}}$ are normalized, but not orthogonal:
$\braket{\ut{p}|\ut{q}}=\delta_{pq}-\p{1-\delta_{pq}}/\p{N-1}$.  We
therefore have
\begin{align}
  \ut{\check{\m h}}_k
  = - \sum_{s\in\set{\pm}} \varepsilon_s \op{\ut{k}^s}
\end{align}
where
\begin{align}
  \varepsilon_\pm \equiv N - 1 \mp 1,
  &&
  \ket{\ut{k}^\pm}
  \equiv \p{1\mp\f1{N-1}}^{-1/2}
  \f{\ket{\ut{0}}\pm\ket{\ut{k}}}{\sqrt{2}}.
\end{align}
The vectors $\ket{\ut{k}^\pm}$ are orthonormal when $k\ne0$.  When
$k=0$, meanwhile, we have
\begin{align}
  \ut{\check{\m h}}_0 = -2\p{N-1} \op{\ut{0}}.
\end{align}
Remembering from \eqref{eq:cycle_vecs} that vectors $\ket{\ut{k}^\pm}$
are associated with the cycle matrices $\m m_{dk}$ defined in
\eqref{eq:cycle} via
\begin{align}
  \ket{\ut{k}^\pm}
  \propto \ket{\ut{0}} \pm \ket{\ut{k}}
  \propto \sum_{d\ne0} \p{1 \pm e^{ik\c d}} \ket{d}
  \to \sum_{d\ne0} \p{1 \pm e^{ik\c d}} \m m_{dk},
\end{align}
one can work out that the vectors $\ket{\ut{k}^\pm}$ correspond to
matrices that are strictly symmetric ($+$) or anti-symmetric ($-$)
under transposition.  The two-body perturbations considered in this
work, however, are defined using coefficient vectors
$\set{\v{\m w}_O}$ that correspond to symmetric matrices
$\set{\m w_O}$, with $w_{Opq}=w_{Oqp}$.  We can therefore neglect the
eigenvectors $\ket{\ut{k}^-}$ and corresponding eigenvalues
$\varepsilon_-$ entirely for the purposes of this work.

Altogether, the excitation energies corresponding to non-trivial
eigenstates of $\ut{\check{\m h}}_k$ are
\begin{align}
  \Delta_0 \equiv -2\p{N-1} - 2 h_0 = 0,
  &&
  \Delta_+ \equiv -\varepsilon_+ - 2 h_0 = N,
\end{align}
while the energy of all orthogonal two-body excitations is
\begin{align}
  \Delta_\perp \equiv - 2 h_0 = 2N - 2.
\end{align}

>>>>

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \bar h_{pq} w_{pq}
  + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  &&
  \bar h_{pq} \equiv 2h_{pq} - h_p - h_q,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
which can be written in matrix form as
\begin{align}
  \bar{\m h}*\m w + \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h}
  = \Delta \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$, $\bar{\m h}$, and $\m w$ are matrices of all $h_{pq}$,
$\bar h_{pq}$ and $w_{pq}$; $\bar{\m h}*\m w$ denotes an element-wise
(Kronecker) product of $\bar{\m h}$ and $\m w$, i.e.~with matrix
elements $\bar h_{pq}w_{pq}$; and
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices.  In order to find solutions to \eqref{eq:cond_2_ref} when
the interaction Hamiltonian $H_0$ is translationally invariant, in
which case $h_{pq}$ and $\bar h_{pq}$ depend only on the separation
$\pm\p{p-q}$, we expand all matrices in a basis that the respects
translational symmetry.  To this end, for all lattice displacements
$d\in\ZZ_L^D$ and wavenumbers $k\in\ZZ_L^D\times2\pi/L$ we define the
{\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms
of these cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
and expand
\begin{align}
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
  &&
  \m h = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \bar{\m h} = \sum_d \bar h_{d,0} \m m_{d,0}.
\end{align}
We can then write the condition in \eqref{eq:cond_2_ref} as a new set
of conditions
\begin{align}
  \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{d\ne0} \bar h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k = \Delta \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a fully symmetric
state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a fixed-energy
perturbation $\V_2^\Delta$ is then performed by projecting each
$k$-component of $\v{\m w}$ onto the corresponding $\Delta$-eigenspace
of $\ut{\check{\m h}}_k$.  If the perturbation $\V_2$ is
translationally invariant, then the $k$-components of $\v{\m w}$ for
all $k\ne 0$ all vanish, so we only need to consider the case of
$k=0$.

@@

Here we solve the single- and two-body eigenvalue problems in the case
of a fully collective spin model with uniform couplings:
$h_{pq}=-1+\delta_{pq}$.  In this case, the single-body excitation
energies $\Delta_k$ are
\begin{align}
  \Delta_k
  = - \sum_{d\ne 0} \sp{\cos\p{d\c k}-1}
  = N \p{1-\delta_{k,0}}.
\end{align}
In order to solve the two-body eigenvalue problem, we expand
\begin{align}
  \ut{\check{\m h}}_k
  &= \sum_{d\ne0} \sp{-2+2\p{N-1}} \op{d}
  - \sum_{\substack{c,d\ne0\\c\ne d}} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c} \\
  &= 2\p{N-1} \1_{N-1}
  - \sum_{c,d\ne0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
\end{align}
where $\1_{N-1}\equiv\sum_{d\ne0}\op{d}$ is the $\p{N-1}\times\p{N-1}$
identity matrix, and define the vectors
\begin{align}
  \ket{\ut{p}}
  \equiv \f1{\sqrt{N-1}} \sum_{d\ne 0} e^{i p\c d}\ket{d},
\end{align}
in terms of which
\begin{align}
  \ut{\check{\m h}}_k
  = 2\p{N-1} \1_{N-1} -\p{N-1} \p{\op{\ut{0}} + \op{\ut{k}}}.
\end{align}
The vectors $\ket{\ut{p}}$ are normalized, but not orthogonal:
$\braket{\ut{p}|\ut{q}}=\delta_{pq}-\p{1-\delta_{pq}}/\p{N-1}$.  We
therefore have
\begin{align}
  \ut{\check{\m h}}_k
  = 2\p{N-1} \1_{N-1} - \sum_{s\in\set{\pm}} \varepsilon_s \op{\ut{k}^s}
\end{align}
where
\begin{align}
  \varepsilon_\pm \equiv N - 1 \mp 1,
  &&
  \ket{\ut{k}^\pm}
  \equiv \p{1\mp\f1{N-1}}^{-1/2}
  \f{\ket{\ut{0}}\pm\ket{\ut{k}}}{\sqrt{2}}.
\end{align}
The vectors $\ket{\ut{k}^\pm}$ are orthonormal when $k\ne0$.  When
$k=0$, meanwhile, we have
\begin{align}
  \ut{\check{\m h}}_0 = 2\p{N-1} \p{\1_{N-1} - \op{\ut{0}}}.
\end{align}
Remembering from \eqref{eq:cycle_vecs} that vectors $\ket{\ut{k}^\pm}$
are associated with the cycle matrices $\m m_{dk}$ defined in
\eqref{eq:cycle} via
\begin{align}
  \ket{\ut{k}^\pm}
  \propto \ket{\ut{0}} \pm \ket{\ut{k}}
  \propto \sum_{d\ne0} \p{1 \pm e^{ik\c d}} \ket{d}
  \to \sum_{d\ne0} \p{1 \pm e^{ik\c d}} \m m_{dk},
\end{align}
one can work out that the vectors $\ket{\ut{k}^\pm}$ correspond to
matrices that are strictly symmetric ($+$) or anti-symmetric ($-$)
under transposition.  The two-body perturbations considered in this
work, however, are defined using coefficient vectors
$\set{\v{\m w}_O}$ that correspond to symmetric matrices
$\set{\m w_O}$, with $w_{Opq}=w_{Oqp}$.  We can therefore neglect the
eigenvectors $\ket{\ut{k}^-}$ and corresponding eigenvalues
$\varepsilon_-$ entirely for the purposes of this work.

Altogether, the excitation energies corresponding to non-trivial
eigenstates of $\ut{\check{\m h}}_k$ are
\begin{align}
  \Delta_0 \equiv 0,
  &&
  \Delta_+ \equiv 2\p{N-1} - \varepsilon_+ = N,
\end{align}
while the energy of all orthogonal two-body excitations is
\begin{align}
  \Delta_\perp \equiv 2\p{N-1}.
\end{align}
